SNP Analysis

### 2019-11-21 snp/SNP_Prediction_Analysis.Rmd
...

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(tidyverse)
library(gridExtra)
library(glmnet)
library(pROC)
#library(ggfortify)
library(BarrettsProgressionRisk)
library(kableExtra)

@@

ct = cor.test(sample.info$ASCAT.SCA.ratio, sample.info$SCA.Ratio)
ct$estimate
```

@@

`r nrow(patient.info)` total endoscopies, `r pander(table(patient.info$Status), caption="Split between P and NP")`

@@

ggplot(sm, aes(Pathology, freq, group=Endoscopy, fill=factor(Endoscopy))) + facet_grid(~Status) + 
  geom_col(position = position_dodge()) + ylim(0,1) + labs(title='Pathology frequency per timepoint')
```

@@

# Quick QC 
In the sWGS data progressor samples show a variance of `r round(pv, 3)` and non-prog variance is `r round(npv,3)`

@@

```{r, echo=F, warning=F, message=F}
qcdata = base::merge(qcdata, sample.info[,c('Samplename','Status','Timepoint','SCA.Ratio')], by='Samplename')

@@

# ggplot(subset(qcdata,SampleType == 'BE'), aes(Timepoint, SCA.Ratio, color=PatientID, group=PatientID)) + geom_point() + geom_line() + 
#   facet_wrap(~Status, scales='free_x', ncol=2) + labs(title='SCA Ratio by Timepoint') + theme(legend.position = 'none')

@@

if (file.exists('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata')) {
  load('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata', verbose=T) 
} else {

  mergedSegs = NULL
  mergedArms = NULL
  length(ptdirs)

@@

    segvals = as.data.frame(data.table::fread(list.files(pt, '*wins_tiled.txt', full.names=T)))
    armvals = as.data.frame(data.table::fread(list.files(pt, '*arms_tiled.txt', full.names=T)))

    segvals = segment.matrix(segvals)
    segvals[is.na(segvals)] = mean(segvals,na.rm=T)

    armvals = segment.matrix(armvals)
    armvals[is.na(armvals)] = mean(armvals,na.rm=T)

@@

  save(mergedSegs, mergedArms, file='tmp_seg_pt.Rdata')
}

@@

# Adjust for ploidy
for (i in 1:nrow(mergedSegs)) {
  sample = rownames(mergedSegs)[i]
  x = unlist(strsplit(sample, '_'))
  #ploidy = mean(subset(qcdata, SampleType == 'BE' & PatientID == x[1] & EndoID == x[2])$Ploidy)
  ploidy = subset(qcdata.endo, PatientID == x[1] & EndoID == x[2] & SampleType == 'BE')$Ploidy_mean
  if (grepl('BLD|gastric', sample, ignore.case = T)) ploidy = 2
  mergedSegs[i,] = mergedSegs[i,]-(ploidy-1)
  mergedArms[i,] = mergedArms[i,]-(ploidy-1)
}

@@

```{r rnormals, echo=F, warning=F, message=F, fig.width=10, fig.height=8, eval=T}
randNormal = sample( grep('BLD',rownames(mergedSegs)[nm.rows], value=T), 1)
pt = unlist(strsplit(randNormal, '_'))[1]
segs = subset(segments.list[[pt]], chr %in% c(1:22) & grepl(sub('_', '.*', randNormal),sample))
segs$chr = factor(segs$chr, levels=c(1:22), ordered = T)

lims = range(segs$medLRR, mergedSegs[randNormal,])
# grid.arrange(
#   ggplot(segs) + facet_grid(~chr, space='free_x', scales='free') + ylim(lims) +
#     geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
#     theme(axis.text.x=element_blank() ) + labs(title='ASCAT median LRR', subtitle=pt),
#   ggplot(melt(as.matrix(normalSegs[randNormal,])), aes(Var1, value)) + geom_point() + ylim(lims) +
#     labs(title='CN Adjusted & winsorized') + theme(axis.text.x=element_blank()),
# top='Random BLD or gastric normal', bottom=randNormal)

@@

#lims = range(segs$medLRR, mergedSegs[sub('\\.LogR','',randBE),])
# grid.arrange(
#   ggplot(segs) + facet_grid(~chr, space='free_x', scales='free')  +
#     geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
#     theme(axis.text.x=element_blank() ) + labs(title='Median LRR, ASCAT segments', subtitle=pt), 
#   ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randBE),])), aes(Var1, value)) + geom_point() +
#     theme(axis.text.x=element_blank()) + labs(title='Winsorized LRR, binned'),  
#  ggplot(melt(adjust.cols(mergedSegs[sub('\\.LogR','',randBE),,drop=F], z.mean, z.sd)), aes(Var2, value)) + geom_point() + theme(axis.text.x=element_blank()) + labs(title='Col adjusted LRR, binned'),  
#top='BE', bottom=randBE)

@@

@@

preds = cbind.data.frame(rownames(prob),prob, rr, stringsAsFactors=F)
colnames(preds) = c('Samplename','Prob', 'RR')

preds = preds %>% mutate(Status = as.character(ifelse(Pathology %in% c('HGD','IMD'), 'P', Status)))

preds %>% filter(grepl('BLD|gastric',Level,ignore.case=T)) %>% group_by(Risk, Level) %>% tally %>% spread(Risk,n)

preds %>% filter(!grepl('BLD|gastric',Level,ignore.case=T)) %>% group_by(Status, Risk, Pathology) %>% tally %>% spread(Risk,n)

# offset = log(0.0225)
# preds = preds %>% dplyr::mutate(
#   'Adj. RR'=RR+offset,
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

preds.normal = preds %>% filter(grepl('BLD|gastric', Level, ignore.case=T))

preds.normal %>% group_by(Risk) %>% tally

length(which(preds.normal$Prob > 0.5))/nrow(preds.normal) # FPR
length(which(preds.normal$Prob < 0.5))/nrow(preds.normal)

@@

preds = sample.info %>% mutate(Sample = paste(PatientID,EndoID,sep='_')) %>% left_join(preds, by=c('Sample'='Samplename'))

@@

preds = preds %>% 
  mutate(Risk = ifelse(Prob > 0.5, 'High', 'Low')) %>% 
  mutate(Risk = factor(Risk, levels=c('Low','High'), ordered=T))

preds %>% group_by(Status, Risk) %>% tally %>% spread(Risk, n) %>%
  left_join(preds %>% group_by(Status) %>% tally, by='Status') %>%
  mutate_if(is.numeric, funs( round(./n,2))) %>% 
  dplyr::select(-n)

@@

rocR = pROC::roc(Status ~ Risk, data=preds, auc=T, ci=T, of='thresholds',transpose=T)
roc.plot(rocR) + labs(title='SNP AUC by discretized risks (low/high)')
pROC::coords(rocR, 'best')

@@

A quick look at the two random samples from earlier.  
`r pander(subset(preds.normal, Samplename == randNormal), caption='Random normal')`

The random BE should show a very high probability, but for some reason is missed entirely:
`r pander(subset(preds, Samplename == randBE), caption='Random BE')`

@@

pred.qc = merge(preds,qcdata.endo, by=c('PatientID','EndoID'))

#dim(pred.qc)
#table(pred.qc$Status)

grid.arrange(
  ggplot(pred.qc, aes(ASCAT.SCA.ratio_mean, Prob, color=Status)) + geom_point(),
  ggplot(pred.qc, aes(Ploidy_max, Prob, color=Status)) + geom_point(),
ncol=2)

pander(cor.test(pred.qc$ASCAT.SCA.ratio_mean, pred.qc$Prob))
```

```{r, warning=F, fig.height=8}
prog = subset(pred.qc, Status == 'P')
p1 = ggplot(prog, aes(Prob)) + facet_wrap(~Timepoint, scales='free_y',nrow=1) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on progressors', y='n Samples', x='Unadjusted Probability')

@@

```{r roc, echo=T}
pred.qc = base::merge( pred.qc, patient.info[,c('PatientID', 'Timepoint Code','Pathology')], by.x=c('PatientID', 'EndoID'), by.y=c('PatientID','Timepoint Code') )

@@

```{r echo=F}
pred.qc.ex = subset(pred.qc, SampleType == 'BE' & !exclude & !wgd & !lowsca)
cuts = seq(0,1,0.1)

@@

```{r regression, echo=F, message=F, eval=T}
folds = 10; splits = 5
## ----- All ----- ##
cache.dir = '~/Data/Ellie/Analysis/SNP/no-low/'
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
 message(paste("loading", file))
 load(file, verbose=T)
 model.performance(performance.at.1se, coefs)$plot + labs(title='SNP CV models')
} else {
  warning("Run 'snp-regression.R' first")
}
#save(arrayDf,labels,patient.info, file=paste(cache.dir,'snp-models.Rdata',sep='/'))

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(tidyverse)
library(gridExtra)
library(glmnet)
library(pROC)
#library(ggfortify)
library(BarrettsProgressionRisk)
library(pander)
library(kableExtra)
library(reshape2)

@@

ct = cor.test(sample.info$ASCAT.SCA.ratio, sample.info$SCA.Ratio)
```

@@

`r nrow(patient.info)` total endoscopies, `r patient.info %>% group_by(Status) %>% tally %>% spread(Status,n) %>% kable(caption="Split between P and NP") %>% kable_styling(full_width=F)`

@@

ggplot(sm %>% filter(!Status == 'Normal'), aes(Pathology, freq, group=Endoscopy, fill=factor(Endoscopy))) + facet_grid(~Status) + 
  geom_col(position = position_dodge()) + ylim(0,1) + labs(title='Pathology frequency per timepoint')
```

@@

# Quick QC

@@

```{r, echo=F, warning=F, message=F}
qcdata = left_join(qcdata, sample.info[,c('Samplename','Status','Timepoint','SCA.Ratio')], by='Samplename')

@@

@@

if (file.exists('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata')) {
  load('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata', verbose=T) 
} else {
  ptdirs = list.dirs('~/Data/Reid_SNP/PerPatient', recursive = F)
  mergedSegs = NULL;  mergedArms = NULL
  length(ptdirs)

@@

    segvals = read_tsv(list.files(pt, '*wins_tiled.txt', full.names=T), col_types = cols(.default = col_double(), 'X1'= col_character())) %>% dplyr::rename(Samplename = X1)
    armvals = read_tsv(list.files(pt, '*arms_tiled.txt', full.names=T), col_types = cols(.default = col_double(), 'X1'= col_character())) %>% dplyr::rename(Samplename = X1)
    #samplenames = segvals$Samplename

    segvals = segvals %>% set_rownames(.$Samplename) %>% dplyr::select(-Samplename) %>% as.matrix 
    armvals = armvals %>% set_rownames(.$Samplename) %>% dplyr::select(-Samplename) %>% as.matrix

    #segvals[is.na(segvals)] = mean(segvals,na.rm=T)
    #armvals[is.na(armvals)] = mean(armvals,na.rm=T)

@@

  save(mergedSegs, mergedArms, file='~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata')
}

@@

# Adjust for ploidy -- this was done in the processing
# for (i in 1:nrow(mergedSegs)) {
#   sample = rownames(mergedSegs)[i]
#   x = unlist(strsplit(sample, '_'))
#   #ploidy = mean(subset(qcdata, SampleType == 'BE' & PatientID == x[1] & EndoID == x[2])$Ploidy)
#   ploidy = subset(qcdata.endo, PatientID == x[1] & EndoID == x[2] & SampleType == 'BE')$Ploidy_mean
#   if (grepl('BLD|gastric', sample, ignore.case = T)) ploidy = 2
#   mergedSegs[i,] = mergedSegs[i,]-(ploidy-1)
#   mergedArms[i,] = mergedArms[i,]-(ploidy-1)
# }

@@

```{r rnormals, echo=F, warning=F, message=F, fig.width=10, fig.height=8, eval=T}

adjust.cols<-function(df, means, sds) {
  nms = colnames(df)

  for (n in nms)
    df[,n] = BarrettsProgressionRisk:::unit.var(df[,n]+0.90, means[[n]], sds[[n]])

  return(df)  
}

randNormal = sample( grep('BLD',rownames(mergedSegs)[nm.rows], value=T), 1)
pt = unlist(strsplit(randNormal, '_'))[1]
segs = subset(segments.list[[pt]], chr %in% c(1:22) & grepl(sub('_', '.*', randNormal),sample))
segs$chr = factor(segs$chr, levels=c(1:22), ordered = T)

lims = range(segs$medLRR, mergedSegs[randNormal,])

grid.arrange(
  ggplot(segs) + facet_grid(~chr, space='free_x', scales='free')  +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) +
    theme(axis.text.x=element_blank() ) + labs(title='Median LRR, ASCAT segments', subtitle=pt),

  ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randNormal),])), aes(Var1, value)) + geom_point() +
        theme(axis.text.x=element_blank()) + labs(title='Ploidy adjusted CN, binned'),

   ggplot(melt(t(adjust.cols(mergedSegs, z.mean, z.sd)[randNormal,])), aes(x=Var2, y=value)) + geom_point() + 
    theme(axis.text.x=element_blank()) + labs(title='Col adjusted CN, binned'),
 top='Random BLD or gastric normal', bottom=randNormal)

@@

lims = range(segs$medLRR, mergedSegs[sub('\\.LogR','',randBE),])
grid.arrange(
  ggplot(segs) + facet_grid(~chr, space='free_x', scales='free')  +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) +
    theme(axis.text.x=element_blank() ) + labs(title='Median LRR, ASCAT segments', subtitle=pt),

  ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randBE),])), aes(Var1, value)) + geom_point() +
    theme(axis.text.x=element_blank()) + labs(title='Ploidy adjusted CN, binned'),

   ggplot(melt(t(adjust.cols(mergedSegs, z.mean, z.sd)[randBE,])), aes(x=Var2, y=value)) + geom_point() + 
    theme(axis.text.x=element_blank()) + labs(title='Col adjusted CN, binned'),
top='BE', bottom=randBE)

@@

median(training.segs)
median(mergedSegs)
median(mergedSegs + 0.98)

median(training.arms)
median(mergedArms)
median(mergedArms + 0.98)

mergedSegs = mergedSegs + 0.98
mergedArms = mergedArms + 0.98

@@

preds = bind_cols('Samplename'=rownames(prob),'Prob'=prob[,1],'RR'=rr[,1]) %>%
  mutate(Risk = case_when(
    Prob <= 0.3 ~ 'Low',
    Prob > 0.3 & Prob < 0.5 ~ 'Moderate',
    Prob >= 0.5 ~ 'High' 
))

normals = filter(preds, grepl('BLD|gastric', Samplename, ignore.case=T))
table(normals$Risk)

preds = filter(preds, !grepl('BLD|gastric', Samplename, ignore.case=T))

preds = preds %>% left_join(patient.info %>% dplyr::select(UniqueSampleID, Pathology, Status), by=c('Samplename'='UniqueSampleID')) %>%
   mutate(Status = as.character(ifelse(Pathology %in% c('HGD','IMD'), 'P', Status)))

preds %>% group_by(Status, Risk) %>% tally %>% spread(Risk,n)

hist(preds$Prob)

length(which(normals$Prob > 0.5))/nrow(normals) # FPR
length(which(normals$Prob < 0.5))/nrow(normals)

@@

#preds = sample.info %>% mutate(Sample = paste(PatientID,EndoID,sep='_')) %>% left_join(preds, by=c('Sample'='Samplename'))

@@

preds %>% group_by(Status, Risk) %>% tally %>% spread(Risk, n) %>%
  left_join(preds %>% group_by(Status) %>% tally, by='Status') %>%
  mutate_if(is.numeric, list( round(./n,2))) %>% 
  dplyr::select(-n)

@@

preds = preds %>%
  mutate(LH = ifelse(Prob > 0.5, 'High', 'Low')) %>%
  mutate(LH = factor(LH, levels=c('Low','High'), ordered=T))

rocR = pROC::roc(Status ~ LH, data=preds, auc=T, ci=T, of='thresholds',transpose=T)
roc.plot(rocR) + labs(title='SNP AUC by discretized risks (low/high)')
pROC::coords(rocR, 'best')

@@

A quick look at the two random samples from earlier.  
`r pander(subset(normals, Samplename == randNormal), caption='Random normal')`

The random BE should show a very high probability, and it does...
`r pander(subset(preds, Samplename == randBE), caption='Random BE')`

@@

#pred.qc = left_join(preds, qcdata.endo %>% dplyr::select(-SampleType), by=c('PatientID', 'EndoID', 'Status')) %>% data.frame

#grid.arrange(
#  ggplot(pred.qc, aes(ASCAT.SCA.ratio_mean, Prob, color=Status)) + geom_point(),
 # ggplot(pred.qc, aes(Ploidy_max, Prob, color=Status)) + geom_point(),
#ncol=2)

#pander(cor.test(pred.qc$ASCAT.SCA.ratio_mean, pred.qc$Prob))
```

```{r, warning=F, fig.height=8, eval=F}
prog = subset(pred.qc, Status == 'P')
p1 = ggplot(prog, aes(Prob)) + facet_wrap(~Timepoint, scales='free_y',nrow=1) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on progressors', y='n Samples', x='Unadjusted Probability')

@@

```{r roc, echo=T, eval=F}
pred.qc = base::merge( pred.qc, patient.info[,c('PatientID', 'Timepoint Code','Pathology')], by.x=c('PatientID', 'EndoID'), by.y=c('PatientID','Timepoint Code') )

@@

```{r echo=F, eval=F}
pred.qc.ex = subset(pred.qc, SampleType == 'BE' & !exclude & !wgd & !lowsca)
cuts = seq(0,1,0.1)

@@

```{r regression, echo=F, message=F, eval=F}
folds = 10; splits = 5
## ----- All ----- ##
cache.dir = '~/Data/Ellie/Analysis/SNP/no-low/'
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (file.exists(file)) {
 message(paste("loading", file))
 load(file, verbose=T)
 model.performance(performance.at.1se, coefs)$plot + labs(title='SNP CV models')
} else {
  warning("Run 'snp-regression.R' first")
}
#save(arrayDf,labels,patient.info, file=paste(cache.dir,'snp-models.Rdata',sep='/'))
